---
title: "Mortality PAF calculation"
author: "Ian Powell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mortality PAF calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rpaf)
```

This vignette describes the process to calculate a *Population Attributable Fraction* (PAF) for mortality given data from a cohort study.

# Quickstart

Once the `rpaf` package has been loaded, mortality PAFs and hazard ratios can be generated with two function calls. Here we use the Mini-Finland Health Survey data as an example, and calculate the PAF for mortality when we modify smoking.

```{r}
smoke_data <- gen_data(
  indata = minifhs,
  id_var = "ID",
  ft_breaks = c(0,5,10,15,17),
  death_ind = "DEATH",
  death_time = "DEATH_FT",
  variables = c("B_COHORT", "SEX", "SMOKE"),
  period_factor = "F_PERIOD",
  time_var = "F_TIME"
)
smoke_summ <- mpaf_summary(
  sr_formula = survival::Surv(F_TIME, DEATH) ~ B_COHORT * F_PERIOD + SEX + SMOKE, 
  mpaf_data = smoke_data,
  modifications = list(SMOKE = c("Never", "<30/day", ">=30/day")),
  covar_model = c("SEX", "SMOKE")
)
smoke_summ
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
