---
title: "Mortality PAF calculation"
author: "Ian Powell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mortality PAF calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rpaf)
```

This vignette describes the process to calculate a *Population Attributable Fraction* (PAF) for mortality given data from a cohort study.

# Quickstart

Once the `rpaf` package has been loaded, mortality PAFs and hazard ratios can be generated with two function calls. Here we use the Mini-Finland Health Survey data as an example, and calculate the PAF for mortality when we modify smoking.

```{r}
smoke_data <- gen_data(
  indata = minifhs,
  id_var = "ID",
  ft_breaks = c(0,5,10,15,17),
  primary_ind = "DEATH",
  primary_time = "DEATH_FT",
  variables = c("B_COHORT", "SEX", "SMOKE"),
  period_factor = "F_PERIOD",
  time_var = "F_TIME"
)
smoke_summ <- mpaf_summary(
  sr_formula = survival::Surv(F_TIME, DEATH) ~ 0 + B_COHORT * F_PERIOD + SEX + SMOKE, 
  paf_data = smoke_data,
  modifications = list(SMOKE = c("Never", "<30/day", ">=30/day")),
  hr_out = c("SEX", "SMOKE")
)
smoke_summ
```

Here, one can see
 - summary output of the survival regression performed by `survival::survreg`
 - the size of the cohort and, if applicable, the number of observations removed
 - hazard ratio estimates and their confidence intervals (where reference levels are reported as `NA`)
 - a list of the modifications performed in the PAF calculations
 - PAFs for mortality over various time periods (given in the row names)

## Modifications

It can be tricky to implement the `modifications` argument properly. In its base form, it is a named list (the names of which correspond to the columns to be modified). In each list element, the first item represents the desired final value after modification. So, for example, if we wanted to modify every entry of the `SMOKE` column to `Never`, we would set
```r
...
  modifications = list(SMOKE = "Never"),
...
```

The remaining elements of each list item represent the values we initially want changed, while those not specified are preserved. So, in the example above, the line
```r
...
  modifications = list(SMOKE = c("Never", "<30/day", ">=30/day")),
...
```
indicates that we want all instances where `SMOKE` is `<30/day` or `>=30/day` to be replaced by `Never`, while the (unspecified) level `Former` is left unmodified.

## Further information
The `mpaf_summary` object contains more information than is reported in the standard print function. For example, it includes the standard errors of the reported PAFs (in transformed space, \(\log(1-\text{PAF})\)), and the design frames (both raw and modified). Specific output can be accessed as from any other list in R, as in the following line of code or the tables below:
```{r}
smoke_summ$se_ipaf0
```

```{r results = 'asis', echo = FALSE}
knitr::kable(smoke_summ$design[c(33:34, 78:79, 167:168),], caption = "Sample rows from `smoke_summ$design`")
knitr::kable(smoke_summ$modified[c(33:34, 78:79, 167:168),], caption = "Sample rows from `smoke_summ$modified`")
```

# Details

The `mpaf_summary` function is simply designed to make life simpler for those who want a quick mortality PAF calculation. It combines a number of steps so that the average user doesn't need to worry about them. The first of these is fitting the survival regression using `est_matrix`. The next is calling the mortality PAF calculation function `mpaf_est_paf`. (If a grouping variable is included, which we shall not discuss here, it splits the data and repeats this call for groupwise PAF estimates, storing them appropriately.)

## Model fitting -- `est_matrix`

The `est_matrix` function fits a survival regression using `survival::survreg` on the prepared data. It takes the model formula for this survival regression and the PAF data we have prepared from `gen_data`. It also takes a list of modifications as described above and our covariate model for reporting hazard ratios. So, instead of the summary function above, we could start with

```{r}
smoke_fit <- est_matrix(
  survival::Surv(F_TIME, DEATH) ~ 0 + B_COHORT * F_PERIOD + SEX + SMOKE, 
  paf_response = smoke_data,
  modifications = list(SMOKE = c("Never", "<30/day", ">=30/day")),
  hr_out = c("SEX", "SMOKE")
)
```
One can also here specify the width of the confidence intervals for hazard ratio reporting, which is by default 95%.

## Population calculation -- `mpaf_est_paf`

This function takes our model fit by `est_matrix` and the original data and calculates the PAF we so desire. So, to find the same mortality PAFs as in our example above, we would call

```r
smoke_paf <- mpaf_est_paf(mpaf_fit = smoke_fit, paf_data = smoke_data)
```

Note that this function optionally takes a `newdata` argument, which is useful when we want to use a different prevalence matrix. This argument **is expanded upon in further detail in another vignette**^[Except not yet.]. One can also specify the confidence interval width again for the PAFs, though now it does not need to be the same as the hazard ratio level, as it would be in the case of using `mpaf_summary`.
